Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-06-28 survey/survey.tex
work on ensuring that the new material in section 3 is consistent with its surroundings. Introduce various macros (and fix others) to make writing more convenient.

\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\idv}[3]{#2 \xrightarrow{#1} #3}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
%% \newcommand{\lid}{\textsf{lid}}
\newcommand{\alt}{~|~}
%% \newcommand{\rid}{\textsf{rid}}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\ensuremath{\mathit{unite}_+\mathit{l}}}
\newcommand{\identrp}{\ensuremath{\mathit{uniti}_+\mathit{l}}}
\newcommand{\identlsp}{\ensuremath{\mathit{unite}_+\mathit{r}}}
\newcommand{\identrsp}{\ensuremath{\mathit{uniti}_+\mathit{r}}}
\newcommand{\swapp}{\ensuremath{\mathit{swap}_+}}
\newcommand{\assoclp}{\ensuremath{\mathit{assocl}_+}}
\newcommand{\assocrp}{\ensuremath{\mathit{assocr}_+}}
\newcommand{\identlt}{\ensuremath{\mathit{unite}_*\mathit{l}}}
\newcommand{\identrt}{\ensuremath{\mathit{uniti}_*\mathit{l}}}
\newcommand{\identlst}{\ensuremath{\mathit{unite}_*\mathit{r}}}
\newcommand{\identrst}{\ensuremath{\mathit{uniti}_*\mathit{r}}}
\newcommand{\swapt}{\ensuremath{\mathit{swap}_*}}
\newcommand{\assoclt}{\ensuremath{\mathit{assocl}_*}}
\newcommand{\assocrt}{\ensuremath{\mathit{assocr}_*}}
\newcommand{\absorbr}{\ensuremath{\mathit{absorbr}}}
\newcommand{\absorbl}{\ensuremath{\mathit{absorbl}}}
\newcommand{\factorzr}{\ensuremath{\mathit{factorzr}}}
\newcommand{\factorzl}{\ensuremath{\mathit{factorzl}}}
\newcommand{\dist}{\ensuremath{\mathit{dist}}}
\newcommand{\factor}{\ensuremath{\mathit{factor}}}
\newcommand{\distl}{\ensuremath{\mathit{distl}}}
\newcommand{\factorl}{\ensuremath{\mathit{factorl}}}
%% \newcommand{\distz}{\mathit{absorbr}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

So mere inhabitation falls far short of our goals of being able to
smoothly deform from one type to another. Let us thus analyze the crux
of the ``problem.'' In logic, we have that $\land$ and $\lor$ are both
\emph{idempotent}: this is the property of any binary operation $*$
where $\forall a. a * a = a$. And it should be clear that an
idempotent operations is a \emph{forgetful} operation: its input has
two copies of $a$, but its output, only one. On the type side,
something more subtle happens. Consider $\top \uplus \top$
versus $\top$; the first has exactly \emph{two} proofs of inhabitation
($\texttt{inl tt}$ and $\texttt{inr tt}$) while the second only one
($\texttt{tt}$). These cannot be put in bijective correspondence. Even
though the ``payload'' \texttt{tt} is the same, forgetting
$\texttt{inl}$ (or \texttt{inr}) throws away information -- something
we have expressly disallowed.  Yes, this should remind you of
Maxwell's daemon: even though the data is the same, they are tagged
differently, and these tags are indeed information, and their
information content must be preserved.

@@

Additive fragment:

\[\begin{array}{rlcl}
 \identlp & (\mathit{right} ~v) &\mapsto & v \\
 \identrp & v &\mapsto & \mathit{right} ~v \\
 \identlsp & (\mathit{left} ~v) &\mapsto & v \\
 \identrsp & v &\mapsto & \mathit{left} ~v \\
 \swapp & (\mathit{left} ~v) &\mapsto & \mathit{right} ~v \\
 \swapp & (\mathit{right} ~v) &\mapsto & \mathit{left} ~v \\
 \assoclp & (\mathit{left} ~v_1) &\mapsto & \mathit{left} ~(\mathit{left} ~v_1) \\
 \assoclp & (\mathit{right} ~(\mathit{left} ~v_2)) &\mapsto & \mathit{left} ~(\mathit{right} ~v_2) \\
 \assoclp & (\mathit{right} ~(\mathit{right} ~v_3)) &\mapsto & \mathit{right} ~v_3 \\
 \assocrp & (\mathit{left} ~(\mathit{left} ~v_1)) &\mapsto & \mathit{left} ~v_1 \\
 \assocrp & (\mathit{left} ~(\mathit{right} ~v_2)) &\mapsto & \mathit{right} ~(\mathit{left} ~v_2) \\
 \assocrp & (\mathit{right} ~v_3) &\mapsto & \mathit{right} ~(\mathit{right} ~v_3) \\
 \end{array}\]
%originally: subcode source isomorphisms.tex:504

Multiplicative fragment:

\[\begin{array}{rlcl}
 \identlt & ((), v) &\mapsto & v \\
 \identrt & v &\mapsto & ((), v) \\
 \identlst & (v, ()) &\mapsto & v \\
 \identrst & v &\mapsto & (v, ()) \\
 \swapt & (v_1, v_2) &\mapsto & (v_2, v_1) \\
 \assoclt & (v_1, (v_2, v_3)) &\mapsto & ((v_1, v_2), v_3) \\
 \assocrt & ((v_1, v_2), v_3) &\mapsto & (v_1, (v_2, v_3)) \\
 \absorbr & (v_1, v_2) & \mapsto & v_1 \\
 \end{array}\]
%originally: subcode source isomorphisms.tex:514

@@

$$
\infer{ \idc ~v \mapsto v}{
	 ~
}
\quad
\infer{ (c_1\odot c_2) ~v_1 \mapsto v_2}{
	 c_1 ~v_1 \mapsto v
	&
	 c_2 ~v \mapsto v_2
}
$$
$$
\infer{ (c_1 \oplus c_2) ~(\mathit{left} ~v_1) \mapsto \mathit{left} ~v_2}{
	 c_1 ~v_1 \mapsto v_2
}
\quad
\infer{ (c_1 \oplus c_2) ~(\mathit{right} ~v_1) \mapsto \mathit{right} ~v_2}{
	 c_2 ~v_1 \mapsto v_2
}
$$
$$
\infer{ (c_1 \otimes c_2) ~(v_1, v_2) \mapsto (v_3, v_4)}{
	 c_1 ~v_1 \mapsto v_3
	&
	 c_2 ~v_2 \mapsto v_4
}
$$
%subcode source isomorphisms.tex:546

@@

\jc{WAVEFRONT of consistency restoration}
Also, the semantics of \ensuremath{\ !~c} are defined in terms of the adjoint
of \ensuremath{c} (denoted by \ensuremath{c^\mathit{^{\dagger} }}),
which we formalize below.

\begin{definition}[Adjoints, \ensuremath{c^\mathit{^{\dagger} }}]
 The adjoint of a combinator \ensuremath{c} is defined by \ensuremath{c^\mathit{^{\dagger} }} as follows:

  \begin{itemize}
  \item For primitive isomophisms \ensuremath{c}, \ensuremath{c^\mathit{^{\dagger} }} is given by its
    isomorphic inverse operator from the table in Definition
    \ref{def:primitives-langRev}.

  \item \ensuremath{(c_1 \otimes c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } \otimes c_2^\mathit{^{\dagger} }}

  \item \ensuremath{(c_1 \oplus c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } \oplus c_2^\mathit{^{\dagger} }}

  \item \ensuremath{(c_1\odot c_2)^\mathit{^{\dagger} } = c_2^\mathit{^{\dagger} } \odot c_1^\mathit{^{\dagger} }}. (Note that the
    order of combinators has been reversed).

\begin{definition}[Observational Equivalence. \ensuremath{c_1=c_2}]
\label{def:pi-eqv}
  Combinators \ensuremath{c_1} and \ensuremath{c_2}, where \ensuremath{c_1 : b_1 \leftrightarrow b_2} and
  \ensuremath{c_2:b_1\leftrightarrow b_2}, are observationally equivalent,\ensuremath{c_1=c_2}, iff

  \ensuremath{\forall  ~v_1:b_1, v_2 : b_2. ~~ c_1 ~v_1 \mapsto v_2 \Longleftrightarrow c_2 ~v_1 \mapsto v_2}.
\end{definition}

\begin{proposition}{Canonical Isomorphism.}
  For any type \ensuremath{b} there exists the isomorphism \ensuremath{b \leftrightarrow b^{|b|}}.
  \begin{proof}
    The fact that such an isomorphism exists is evident from the
    definition of size and what it means for two types to be
    isomorphic. While many equivalent constructions are possible
    for any type \ensuremath{b}, one such construction is given by \ensuremath{\lbrack\!\lbrack b\rbrack\!\rbrack}:

\[\begin{array}{rclr}
 \lbrack\!\lbrack 0 \rbrack\!\rbrack & = & \idc \\
 \lbrack\!\lbrack 1 \rbrack\!\rbrack & = & \idc \\
 \lbrack\!\lbrack 1+b \rbrack\!\rbrack & = & \idc \oplus \lbrack\!\lbrack b \rbrack\!\rbrack \\
 \lbrack\!\lbrack (b_1+b_2)+b_3 \rbrack\!\rbrack & = & \assocrp \odot \lbrack\!\lbrack b_1+(b_2+b_3) \rbrack\!\rbrack \\
 \lbrack\!\lbrack b_1+b_2 \rbrack\!\rbrack & = & ( \lbrack\!\lbrack b_1 \rbrack\!\rbrack \oplus \idc) \odot \lbrack\!\lbrack b_1^{|b_1|} + b_2 \rbrack\!\rbrack \\
 \lbrack\!\lbrack (b_1\times b_2)\times b_3 \rbrack\!\rbrack & = & \assocrt \odot \lbrack\!\lbrack b_1\times (b_2\times b_3) \rbrack\!\rbrack \\
 \lbrack\!\lbrack (b_1+b_2)\times b_3 \rbrack\!\rbrack & = & \dist \odot \lbrack\!\lbrack b_1\times b_3+b_2\times b_3 \rbrack\!\rbrack \\
 \end{array}\]
%subcode source isomorphisms.tex:598

@@

Combinators of \ensuremath{\Pi } can be written in terms of the operators
described in definition \ref{def:langRev} or via a graphical language
similar in spirit to those developed for Geometry of Interaction
\cite{DBLP:conf/popl/Mackie95} and string diagrams for category theory
\cite{selinger-graphical}. Modulo some conventions and shorthand we
describe here, the wiring diagrams are equivalent to the operator
based (syntactic) description of programs.  \ensuremath{\Pi } combinators
expressed in this graphical language look like ``wiring
diagrams''. Values take the form of ``particles'' that flow along the
wires. Computation is expressed by the flow of particles.

@@

\item
Associativity is entirely skipped in the graphical language. Hence
three parallel wires may be inferred as \ensuremath{b_1\times (b_2\times b_3)} or
\ensuremath{(b_1\times b_2)\times b_3}, based on the context. This is much like handling of
associativity in the graphical representations of categories.

@@

\item
The morphisms that witness that \ensuremath{0} and \ensuremath{1} are the additive and
  multiplicative units are represented as shown below. Note that since there
  is no value of type 0, there can be no particle on a wire of type \ensuremath{0}.
  Also since the monoidal units can be freely introduced and eliminated, in
  many diagrams they are omitted and dealt with explicitly only when they are
  of special interest.

@@

\ensuremath{c : b \times (1+1) \leftrightarrow b + b}

\ensuremath{c = \swapt \odot \dist \odot (\identlt + \identlt)}

\begin{center}
\scalebox{1.5}{
%%subcode-line{pdfimage}[diagrams/example1-crop.pdf]
\includegraphics{inkscape/example1.pdf}
}
\end{center}

The textual representation of $\Pi$ is quite cumbersome: the
underlying principles however become quite intuitive when represented
in a more suggestive visual notation. There exist various graphical
notations for the monoidal categories that model $\Pi$ and related
languages including Penrose diagrams~\cite{selinger-graphical} and
Geometry of Interaction (GoI) machines and Proof
Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In this section, we
propose a novel graphical representation of $\Pi$ types and
combinators drawn from the resource usage embedded in the operational
semantics. As suggested by the operational semantics, the type $A * B$
can be viewed as denoting \emph{spatial composition}, where
\emph{both} $A$ and $B$ must happen in \emph{some} order or perhaps
simultaneously, and where $A + B$ denotes \emph{temporal composition},
where exactly \emph{one of} $A$ and $B$ must happen at a given time.
\jc{I think the temporal aspect is stretching things too far. Could we
call it ``choice composition''? (Or something better). To me spatial
composition somehow seems like sequential composition (i.e. composition
in time), which is definitely not what we want. I think the informal
description (both although perhaps concurrently, and exactly one of)
is better. In section 3 we called them sequential, choice and
parallel. We should be consistent.}

\jc{An additional ``problem'' is that since our circuits have a
left-to-right orientation, $A * B$ drawn as boxes left-to-right
gives an implicit ordering that is going to be hard to ignore.
I'm starting to think that the diagrams in the MO answer might
be better: use a colour (or like in Selinger's paper, at the
very end), symbols in the nodes that split.  So both would
be displayed as vertical stacks (only). Because I think that
swap* is too hard to justify in this notation. Of course,
if we could do things in 3D and $*$ was displayed using
(say) depth, that would be different again.}

@@

Other than distributity, $\Pi$ combinators simply allow us to shuffle
the components of a stack around as well as shuffle the positions of
components in the plane.

% \amr{
%   Let’s have a geometric representation of spaces:
% - A + B means putting A next to B
% - A * B means putting A and B on two perpendicular axes and filling
% the square (volume) between them
% All the combinators would indeed have a natural action and if we have
% a few nice pictures it would be great. What do you think? }

>>>>

\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
%% \newcommand{\lid}{\textsf{lid}}
\newcommand{\alt}{~|~}
%% \newcommand{\rid}{\textsf{rid}}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\ensuremath{\mathit{unite}_+\mathit{l}}}
\newcommand{\identrp}{\ensuremath{\mathit{uniti}_+\mathit{l}}}
\newcommand{\identlsp}{\ensuremath{\mathit{unite}_+\mathit{r}}}
\newcommand{\identrsp}{\ensuremath{\mathit{uniti}_+\mathit{r}}}
\newcommand{\swapp}{\ensuremath{\mathit{swap}_+}}
\newcommand{\assoclp}{\ensuremath{\mathit{assocl}_+}}
\newcommand{\assocrp}{\ensuremath{\mathit{assocr}_+}}
\newcommand{\identlt}{\ensuremath{\mathit{unite}_*\mathit{l}}}
\newcommand{\identrt}{\ensuremath{\mathit{uniti}_*\mathit{l}}}
\newcommand{\identlst}{\ensuremath{\mathit{unite}_*\mathit{r}}}
\newcommand{\identrst}{\ensuremath{\mathit{uniti}_*\mathit{r}}}
\newcommand{\swapt}{\ensuremath{\mathit{swap}_*}}
\newcommand{\assoclt}{\ensuremath{\mathit{assocl}_*}}
\newcommand{\assocrt}{\ensuremath{\mathit{assocr}_*}}
\newcommand{\absorbr}{\ensuremath{\mathit{absorbr}}}
\newcommand{\absorbl}{\ensuremath{\mathit{absorbl}}}
\newcommand{\factorzr}{\ensuremath{\mathit{factorzr}}}
\newcommand{\factorzl}{\ensuremath{\mathit{factorzl}}}
\newcommand{\dist}{\ensuremath{\mathit{dist}}}
\newcommand{\factor}{\ensuremath{\mathit{factor}}}
\newcommand{\distl}{\ensuremath{\mathit{distl}}}
\newcommand{\factorl}{\ensuremath{\mathit{factorl}}}
%% \newcommand{\distz}{\mathit{absorbr}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\sem}[1]{\ensuremath{\llbracket{#1}\rrbracket}}

@@

So mere inhabitation falls far short of our goals of being able to
smoothly deform from one type to another. Let us thus analyze the crux
of the ``problem.'' In logic, we have that $\land$ and $\lor$ are both
\emph{idempotent}: this is the property of any binary operation $*$
where $\forall a. a * a = a$. And it should be clear that an
idempotent operations is a \emph{forgetful} operation: its input has
two copies of $a$, but its output, only one. On the type side,
something more subtle happens. Consider $\top \uplus \top$
versus $\top$; the first has exactly \emph{two} proofs of inhabitation
($\inl{\texttt{tt}}$ and $\inr{\texttt{tt}}$) while the second only one
($\texttt{tt}$). These cannot be put in bijective correspondence. Even
though the ``payload'' \texttt{tt} is the same, forgetting
$\texttt{inl}$ (or \texttt{inr}) throws away information -- something
we have expressly disallowed.  Yes, this should remind you of
Maxwell's daemon: even though the data is the same, they are tagged
differently, and these tags are indeed information, and their
information content must be preserved.

@@

Identity:
\[\begin{array}{rlcl}
 \idc & v &\mapsto & v \\
 \end{array}\]

Additive fragment:
\[\begin{array}{rlcl}
 \identlp & (\mathit{right} ~v) &\mapsto & v \\
 \identrp & v &\mapsto & \mathit{right} ~v \\
 \identlsp & (\mathit{left} ~v) &\mapsto & v \\
 \identrsp & v &\mapsto & \mathit{left} ~v \\
 \swapp & (\mathit{left} ~v) &\mapsto & \mathit{right} ~v \\
 \swapp & (\mathit{right} ~v) &\mapsto & \mathit{left} ~v \\
 \assoclp & (\mathit{left} ~v_1) &\mapsto & \mathit{left} ~(\mathit{left} ~v_1) \\
 \assoclp & (\mathit{right} ~(\mathit{left} ~v_2)) &\mapsto & \mathit{left} ~(\mathit{right} ~v_2) \\
 \assoclp & (\mathit{right} ~(\mathit{right} ~v_3)) &\mapsto & \mathit{right} ~v_3 \\
 \assocrp & (\mathit{left} ~(\mathit{left} ~v_1)) &\mapsto & \mathit{left} ~v_1 \\
 \assocrp & (\mathit{left} ~(\mathit{right} ~v_2)) &\mapsto & \mathit{right} ~(\mathit{left} ~v_2) \\
 \assocrp & (\mathit{right} ~v_3) &\mapsto & \mathit{right} ~(\mathit{right} ~v_3) \\
 \end{array}\]
%originally: subcode source isomorphisms.tex:504

Multiplicative fragment:
\[\begin{array}{rlcl}
 \identlt & ((), v) &\mapsto & v \\
 \identrt & v &\mapsto & ((), v) \\
 \identlst & (v, ()) &\mapsto & v \\
 \identrst & v &\mapsto & (v, ()) \\
 \swapt & (v_1, v_2) &\mapsto & (v_2, v_1) \\
 \assoclt & (v_1, (v_2, v_3)) &\mapsto & ((v_1, v_2), v_3) \\
 \assocrt & ((v_1, v_2), v_3) &\mapsto & (v_1, (v_2, v_3)) \\
 \absorbr & (v_1, v_2) & \mapsto & v_1 \\
 \end{array}\]
%originally: subcode source isomorphisms.tex:514

@@

$$
\infer{ (c_1\odot c_2) ~v_1 \mapsto v_2}{
	 c_1 ~v_1 \mapsto v
	&
	 c_2 ~v \mapsto v_2
}
$$
$$
\infer{ (c_1 \oplus c_2) ~(\mathit{left} ~v_1) \mapsto \mathit{left} ~v_2}{
	 c_1 ~v_1 \mapsto v_2
}
\quad
\infer{ (c_1 \oplus c_2) ~(\mathit{right} ~v_1) \mapsto \mathit{right} ~v_2}{
	 c_2 ~v_1 \mapsto v_2
}
$$
$$
\infer{ (c_1 \otimes c_2) ~(v_1, v_2) \mapsto (v_3, v_4)}{
	 c_1 ~v_1 \mapsto v_3
	&
	 c_2 ~v_2 \mapsto v_4
}
$$
%subcode source isomorphisms.tex:546

@@

As we mentioned before, $!$ is a defined combinator.

\begin{definition}[Adjoint, \ensuremath{!~ c}]
 The adjoint of a combinator \ensuremath{c} is defined as follows:

  \begin{itemize}
  \item For primitive isomophisms \ensuremath{c}, \ensuremath{!~ c} is given by its
    inverse from Figures~\ref{pi-terms} and~\ref{more-pi}.

  \item \ensuremath{!(c_1 \otimes c_2) =\ !c_1 \otimes~ !c_2}

  \item \ensuremath{!(c_1 \oplus c_2) =\ !c_1 \oplus~ !c_2}

  \item \ensuremath{!(c_1\odot c_2) =\ !c_2 \odot~ !c_1}. (Note that the
    order of combinators has been reversed).

  \end{itemize}
\end{definition}

\noindent We can further define that two combinators are 
\emph{observationally equivalent} if only all values of their
common domain, they evaluate to identical values.  More precisely,
we will say that for combinators $c_1, c_2 : b_1 \leftrightarrow b_2$,
$c_1~=~c_2$ whenever
\[
  \forall  ~v_1:b_1, v_2 : b_2. ~~ c_1 ~v_1 \mapsto v_2 \text{if and only if\ } c_2 ~v_1 \mapsto v_2
\]

Previously we had seen that types $b$ correspond to natural numbers,
more precisely~$|b|$, the size of $b$.  We had previously established that 
for any natural number $n$, there is a canonical set of size $n$,
which we denoted $[n]$. Furthermore, we can also define a
canonical \emph{type} that size, which we will denote $\sharp b$,
i.e. $\sharp b$ is a canonical type of size $|b|$.

\begin{definition}($\sharp$). By recursion on $|b|$.  First define a
type associated to each size. We will revert to using type notation
for greater clarity of this definition:
\[\begin{array}{rcl}
  \tau~ 0 & = & \bot \\
  \tau~ (1 + n) & = & \top \uplus \tau~ n \\
 \end{array}\]
\noindent so that we can define $\sharp b = \tau~ |b|$.
\end{definition}

We are now ready to go further and establish
that there is always an equivalence between.

\begin{proposition}
  For any type \ensuremath{b} there exists an isomorphism \ensuremath{b \leftrightarrow \sharp b}.
  \begin{proof}
    The fact that such an isomorphism exists is evident from the
    definition of size and what it means for two types to be
    isomorphic. While many equivalent constructions are possible
    for any type \ensuremath{b}, one such construction is given by 
    \sem{b}:

\[\begin{array}{rclr}
 \sem{0} & = & \idc \\
 \sem{1} & = & \idc \\
 \sem{1+b} & = & \idc \oplus \sem{b} \\
 \sem{(b_1+b_2)+b_3} & = & \assocrp \odot \sem{b_1 + (b_2 + b_3)} \\
 \sem{b_1 + b_2} & = & (\sem{b_1} \oplus \idc) \odot \sem{ \sharp b_1 + b_2 } \\
 \sem{0 \times b_2} & = & \absorbr \\
 \sem{1 \times b_2} & = & \identlt \odot \sem{b_2} \\
 \sem{(b_1\times b_2)\times b_3} & = & \assocrt \odot \sem{b_1\times (b_2\times b_3)} \\
 \sem{(b_1+b_2)\times b_3} & = & \dist \odot \sem{b_1\times b_3+b_2\times b_3} \\
 \end{array}\]
%originally subcode source isomorphisms.tex:598

@@

Combinators of \ensuremath{\Pi } can be written in terms of the operators
described previously or via a graphical language
similar in spirit to those developed for Geometry of Interaction
\cite{DBLP:conf/popl/Mackie95} and string diagrams for category theory
\cite{selinger-graphical}. Modulo some conventions and shorthand we
describe here, the wiring diagrams are equivalent to the operator
based (syntactic) description of programs.  \ensuremath{\Pi } combinators
expressed in this graphical language look like ``wiring
diagrams''. Values take the form of ``particles'' that flow along the
wires. Computation is expressed by the flow of particles.

@@

\item
Associativity is entirely skipped in the graphical language. Hence
three parallel wires may be inferred as \ensuremath{b_1\times (b_2\times b_3)} or
\ensuremath{(b_1\times b_2)\times b_3}, based on the context. This is much like handling of
associativity in the graphical representations of categories as well as
that for monoidal categories.

@@

\item
The morphisms that witness that \ensuremath{0} and \ensuremath{1} are the additive and
  multiplicative units are represented as shown below. Note that since there
  is no value of type 0, there can be no particle on a wire of type \ensuremath{0}.
  Also since the monoidal units can be freely introduced and eliminated, sometimes
  they are omitted.  However, as this is in fact dangerous, as explained by
  many diagrams they are omitted and dealt with explicitly only when they are
  of special interest.

@@

\ensuremath{c : b \times (1+1) \leftrightarrow b + b}
\ensuremath{c = \swapt \odot \dist \odot (\identlt + \identlt)}
\begin{center}
\scalebox{1.5}{
%%subcode-line{pdfimage}[diagrams/example1-crop.pdf]
\includegraphics{inkscape/example1.pdf}
}
\end{center}

\jc{WAVEFRONT of consistency pass}
The textual representation of $\Pi$ is quite cumbersome: the
underlying principles however become quite intuitive when represented
in a more suggestive visual notation. There exist various graphical
notations for the monoidal categories that model $\Pi$ and related
languages including Penrose diagrams~\cite{selinger-graphical} and
Geometry of Interaction (GoI) machines and Proof
Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In this section, we
propose a novel graphical representation of $\Pi$ types and
combinators drawn from the resource usage embedded in the operational
semantics. As suggested by the operational semantics, the type $A * B$
can be viewed as denoting \emph{spatial composition}, where
\emph{both} $A$ and $B$ must happen in \emph{some} order or perhaps
simultaneously, and where $A + B$ denotes \emph{temporal composition},
where exactly \emph{one of} $A$ and $B$ must happen at a given time.
\jc{I think the temporal aspect is stretching things too far. Could we
call it ``choice composition''? (Or something better). To me spatial
composition somehow seems like sequential composition (i.e. composition
in time), which is definitely not what we want. I think the informal
description (both although perhaps concurrently, and exactly one of)
is better. In section 3 we called them sequential, choice and
parallel. We should be consistent.}

\begin{comment}
\jc{An additional ``problem'' is that since our circuits have a
left-to-right orientation, $A * B$ drawn as boxes left-to-right
gives an implicit ordering that is going to be hard to ignore.
I'm starting to think that the diagrams in the MO answer might
be better: use a colour (or like in Selinger's paper, at the
very end), symbols in the nodes that split.  So both would
be displayed as vertical stacks (only). Because I think that
swap* is too hard to justify in this notation. Of course,
if we could do things in 3D and $*$ was displayed using
(say) depth, that would be different again.}

@@

Other than distributity, $\Pi$ combinators simply allow us to shuffle
the components of a stack around as well as shuffle the positions of
components in the plane.
\end{comment}
% \amr{
%   Let’s have a geometric representation of spaces:
% - A + B means putting A next to B
% - A * B means putting A and B on two perpendicular axes and filling
% the square (volume) between them
% All the combinators would indeed have a natural action and if we have
% a few nice pictures it would be great. What do you think? }
